% !TEX root = ../main.tex
\chapter{System-on-Chip Design}
\label{ch:platform}

This chapter describes the system that hosts the SpMM accelerator: what the
SoC looks like as a whole, why a soft RISC-V core was chosen as the host, how
memory and the bus are organised, and how firmware drives the accelerator
through a small bank of memory-mapped registers. The accelerator itself is
treated as a black box here; its internal datapath and optimisation are the
subject of Chapter~\ref{ch:design}.

\section{System Overview}

The SoC is synthesised entirely into FPGA fabric on the Terasic DE1-SoC
development board~\citep{de1soc_manual}. The hard ARM Cortex-A9 on the HPS side
of the device is unused, so the whole design is self-contained in
programmable logic. All benchmark code and data reside on chip: firmware,
CSR arrays, the dense matrix $B$, and the output matrix $C$ all reside in a
shared 64\,KB dual-port block RAM. The entire SoC is clocked from the DE1-SoC's
50 MHz on-board oscillator.

Keeping the design entirely in FPGA fabric removes the HPS, Linux, AXI
bridge, and external-memory effects from the measurement path. This makes
CPU and accelerator cycle counts directly comparable, but limits benchmark
sizes to what fits in the 64\,KB shared BRAM. The shared-memory structure
is therefore the central platform constraint.

Figure~\ref{fig:soc-block} shows the top-level block diagram. The PicoRV32
CPU drives a valid/ready memory bus into the SoC interconnect, which decodes
the address and routes to four slave regions: BRAM Port~A, the SpMM
accelerator's MMIO registers, GPIO, and the UART. The accelerator
additionally accesses BRAM directly through Port~B, which lets it stream
operands and write back concurrently with CPU activity on Port~A. All blocks
are clocked from the 50\,MHz on-board oscillator.

\begin{figure}[htbp]
  \centering
  \begin{tikzpicture}[
    font=\small,
    block/.style={draw, thick, rounded corners=2pt, align=center},
    cpu/.style={block, fill=blue!12, minimum width=2.6cm, minimum height=1.4cm},
    accel/.style={block, fill=orange!15, minimum width=2.8cm, minimum height=1.4cm},
    intc/.style={block, fill=gray!15, minimum width=11.0cm, minimum height=0.9cm},
    ram/.style={block, fill=purple!10, minimum width=3.6cm, minimum height=1.4cm},
    periph/.style={block, font=\footnotesize, minimum width=2.0cm, minimum height=1.2cm},
    clk/.style={block, fill=yellow!25, minimum width=4.6cm, minimum height=0.6cm,
                font=\footnotesize},
    arr/.style={-{Stealth[length=2.5mm]}, thick},
    biarr/.style={{Stealth[length=2mm]}-{Stealth[length=2mm]}, thick},
    portlbl/.style={font=\scriptsize, fill=white, inner sep=1.5pt}
  ]
    \node[clk] (clk) at (0, 5.0)
      {50\,MHz on-board oscillator (\texttt{clk\_50mhz})};

    \node[cpu]   (cpu) at (-4.0, 3.3) {PicoRV32 CPU\\\footnotesize RV32IM};
    \node[accel] (acc) at ( 4.0, 3.3) {SpMM Accelerator\\\footnotesize \texttt{accel\_top}};

    \node[intc] (bus) at (0, 1.5)
      {SoC Interconnect \quad (address decode)};

    \node[ram] (ram) at (-4.0, -0.6)
      {64\,KB shared\\dual-port BRAM\\\footnotesize\texttt{altsyncram}};
    \node[periph, fill=green!12] (gpio) at (0.0, -0.6)
      {GPIO\\\scriptsize LEDs/btns\\\scriptsize 7-seg};
    \node[periph, fill=cyan!15]  (uart) at (4.0, -0.6)
      {UART\\\scriptsize\texttt{simpleuart}\\\scriptsize 115200 baud};

    \draw[arr, densely dashed] (clk.south) -- (0, 4.3)
      node[portlbl, midway, right=2pt]{system clock};
    \draw[densely dashed, thick] (-4.0, 4.3) -- (4.0, 4.3);
    \draw[arr, densely dashed] (-4.0, 4.3) -- (cpu.north);
    \draw[arr, densely dashed] ( 4.0, 4.3) -- (acc.north);
    \draw[arr, densely dashed] ( 0.0, 4.3) -- (bus.north);

    \draw[biarr] (cpu.south) -- (cpu.south |- bus.north)
      node[portlbl, midway, right=2pt]{valid/ready};
    \draw[biarr] (acc.south) -- (acc.south |- bus.north)
      node[portlbl, midway, right=2pt]{MMIO};

    \draw[biarr] (bus.south -| ram.north) -- (ram.north)
      node[portlbl, midway, right=2pt]{Port~A};
    \draw[biarr] (bus.south -| gpio.north) -- (gpio.north);
    \draw[biarr] (bus.south -| uart.north) -- (uart.north);

    \draw[biarr] (acc.east) -- ++(0.9, 0) -- ++(0, -5.0) -| (ram.south);
    \node[portlbl] at (6.5, 1.5) {Port~B (direct)};
  \end{tikzpicture}
  \caption{Top-level RISC-V SoC block diagram.}
  \label{fig:soc-block}
\end{figure}

\section{Why a Soft RISC-V SoC}

The host processor is PicoRV32~\citep{picorv32}, a compact RV32IM soft-core
implementing the RISC-V base integer instruction set with the hardware integer
multiply extension~\citep{riscv_isa}. Three practical reasons drove the
choice.

The compact RV32IM core leaves the majority of the Cyclone~V fabric
available for the accelerator and the 64\,KB BRAM; the native
\texttt{rdcycle} performance counter CSR makes all benchmark timing
cycle-accurate without external instrumentation; and the plain
valid/ready memory bus drops onto the SoC interconnect with no AXI
adapter. Table~\ref{tab:picorv32-config} lists the full PicoRV32
configuration applied in this implementation.

\begin{table}[htbp]
  \centering
  \small
  \begin{tabular}{p{3.4cm}p{3.0cm}p{7.5cm}}
  \toprule
  Parameter & Value & Rationale \\
  \midrule
  Base ISA              & RV32I & Minimum integer ISA targeted by the GCC toolchain. \\[2pt]
  M extension           & Enabled & Hardware \texttt{MUL}/\texttt{MULH} so the software SpMM baseline is not penalised by emulated multiply. \\[2pt]
  C extension           & Disabled & Compressed ISA off; firmware compiled for aligned RV32IM only. \\[2pt]
  \texttt{rdcycle} / counters & Enabled & 32-bit cycle counter CSR drives all benchmark timing without external instrumentation. \\[2pt]
  Interrupts            & Disabled & Bare-metal harness uses MMIO polling; no IRQ controller is present in the SoC. \\[2pt]
  Reset PC              & \texttt{0x0000\_0000} & Entry point of \texttt{start.S} at the base of the on-chip RAM. \\[2pt]
  Stack pointer (reset) & \texttt{0x0000\_FF00} & PicoRV32 reset value; \texttt{start.S} then loads \texttt{sp} from \texttt{\_\_stack\_top} before \texttt{main}. \\[2pt]
  Memory interface      & Native valid/ready & 32-bit address/data with 4-bit \texttt{wstrb}; drops onto the SoC interconnect with no AXI bridge. \\
  \bottomrule
  \end{tabular}
  \caption{PicoRV32 configuration applied in the implemented SoC,
           recorded from the synthesisable wrapper.}
  \label{tab:picorv32-config}
\end{table}

The main limitation is that PicoRV32 is an in-order, non-pipelined core
with no cache, so the software baseline is weaker than a modern
superscalar or vector CPU. The comparison is still appropriate because
both software and hardware run on the same FPGA SoC and use the same
memory system. The M extension is enabled so the software reference uses
hardware multiply rather than an artificially slow software routine.

\section{Target FPGA Platform}

The Cyclone~V 5CSEMA5F31C6 on the DE1-SoC provides the resources summarised
in Table~\ref{tab:fpga-resources}~\citep{cyclonev_handbook}.

\begin{table}[htbp]
  \centering
  \small
  \begin{tabular}{lrr}
  \toprule
  Resource & Total available & Used (final post-fit) \\
  \midrule
  Adaptive Logic Modules (ALMs) & 32,070 & 15,745 (49\%) \\
  Logic registers & 128,280 & 20,569 (16\%) \\
  Block memory bits & 4,065,280 & 526,336 (13\%) \\
  DSP 18$\times$18 multiplier blocks & 87 & 21 (24\%) \\
  I/O pins & 457 & 69 (15\%) \\
  \bottomrule
  \end{tabular}
  \caption{Cyclone~V 5CSEMA5F31C6 resource availability and final
           post-fit design utilisation for the complete \texttt{soc\_top}.}
  \label{tab:fpga-resources}
\end{table}

The critical resources are M10K BRAMs and DSP blocks. The shared 64\,KB
RAM consumes the bulk of the block memory bits~\citep{intel_m10k}, while
the eight MAC lanes, the DSP-based accelerator controller paths, and the
PicoRV32 hardware multiplier together use 21 DSP
blocks~\citep{intel_dsp_blocks}. The final utilisation leaves roughly
half the logic fabric and three-quarters of the DSP blocks free for
further extension.

\section{Memory and Interconnect Architecture}
\label{sec:memory-arch}

\subsection{Bus Protocol}

The SoC interconnect uses the native PicoRV32 valid/ready memory bus with
32-bit address and data signals, a 4-bit byte-enable mask, and a
read/write indicator. All signals are synchronous to the 50\,MHz system
clock. BRAM reads have one registered cycle of latency, while MMIO
register accesses are combinatorial.

A read transaction proceeds as follows. On cycle 0 the CPU asserts
\texttt{mem\_valid} together with \texttt{mem\_addr}, drives
\texttt{mem\_wstrb}~$=\texttt{0000}$ and \texttt{mem\_wdata}~$=0$. The
interconnect decodes the address, selects BRAM Port~A, and forwards
the address to the registered RAM. On cycle 1 the BRAM presents the
registered word on its data output; the interconnect drives that word
onto \texttt{mem\_rdata} and asserts \texttt{mem\_ready} for one cycle.
The CPU latches \texttt{mem\_rdata} on the same edge, deasserts
\texttt{mem\_valid}, and continues. A read therefore completes in two
cycles end-to-end: one address phase and one registered BRAM read.

\subsection{Dual-Port BRAM Organisation}

The 64\,KB on-chip RAM is implemented as an Altera \texttt{altsyncram}
true-dual-port memory. Port~A is used by the CPU with byte enables, while
Port~B is used by the accelerator with word-granular accesses.
Figure~\ref{fig:bram-dualport} shows the arrangement.

\begin{figure}[htbp]
  \centering
  \begin{tikzpicture}[
    font=\small,
    box/.style={draw, thick, rounded corners=2pt, minimum height=1.2cm,
                minimum width=2.6cm, align=center},
    memregion/.style={draw, thick, minimum width=5.2cm, minimum height=0.85cm,
                      align=center, font=\footnotesize},
    portlabel/.style={font=\footnotesize\itshape, inner sep=1pt},
    arr/.style={-{Stealth[length=2.5mm]}, thick}
  ]
    \node[memregion, fill=gray!10] (fw)   at (0,  2.4) {Firmware + stack};
    \node[memregion, fill=gray!10] (csr)  at (0,  1.3) {CSR arrays (rowptr, colidx, values)};
    \node[memregion, fill=gray!10] (bmat) at (0,  0.2) {Dense matrix $B$};
    \node[memregion, fill=gray!10] (cmat) at (0, -0.9) {Output matrix $C$};

    \node[draw, thick, fit=(fw) (csr) (bmat) (cmat),
          inner sep=7pt, rounded corners=3pt,
          label={[font=\small\bfseries, yshift=2pt]above:%
                 64\,KB shared BRAM (\texttt{altsyncram})}] (bram) {};

    \node[box, fill=blue!10]   (cpu) at (-6.4, 0.75) {PicoRV32\\CPU};
    \node[box, fill=orange!15] (acc) at ( 6.4, 0.75) {SpMM\\Accelerator};

    \draw[arr] (cpu.east) -- (bram.west |- cpu.east)
      node[portlabel, above, midway]{Port~A}
      node[portlabel, below, midway]{byte-granular};
    \draw[arr] (acc.west) -- (bram.east |- acc.west)
      node[portlabel, above, midway]{Port~B}
      node[portlabel, below, midway]{word-granular};
  \end{tikzpicture}
  
  \caption{Dual-port BRAM organisation. The CPU and accelerator drive
           independent ports on the same memory array, so they can read or
           write different addresses on the same cycle without arbitration.
           }
  \label{fig:bram-dualport}
\end{figure}

Because both ports address the same underlying array, the firmware
contract treats access ownership as a function of accelerator state.
Before \texttt{CTRL.start} is pulsed, the CPU owns the buffers and
stages the CSR arrays (\texttt{rowptr}, \texttt{colidx},
\texttt{values}), the dense matrix $B$, and a cleared output region
$C$. While \texttt{STATUS.busy} is asserted, the accelerator is the
sole reader of the CSR arrays and $B$ on Port~B and the sole writer of
$C$; the CPU restricts itself to MMIO polling and does not touch any
accelerator-owned word. Ownership returns to the CPU only after
\texttt{STATUS.done} is asserted and acknowledged. Because the firmware
never schedules a Port~A access to the same word that the accelerator
is reading or writing on Port~B, the design does not rely on the
device-specific dual-port collision behaviour of \texttt{altsyncram},
which is undefined for simultaneous same-address access on different
ports.

\subsection{Address Space}

Table~\ref{tab:system-memory-map} summarises the full memory map.

\begin{table}[htbp]
  \centering
  \small
  \begin{tabular}{lp{3.0cm}p{5.5cm}}
  \toprule
  Address range & Peripheral & Purpose \\
  \midrule
  \texttt{0x0000\_0000--0x0000\_FFFF} & On-chip RAM (64\,KB) &
    Firmware code, stack, CSR arrays (\texttt{rowptr}, \texttt{colidx},
    \texttt{values}), dense matrix $B$, output matrix $C$ \\[4pt]
  \texttt{0x1000\_0000--0x1000\_00FF} & SpMM accelerator MMIO &
    Configuration registers, control, and status \\[4pt]
  \texttt{0x2000\_0000--0x2000\_000F} & GPIO &
    LED output, switch and button input, seven-segment display \\[4pt]
  \texttt{0x2000\_0100--0x2000\_01FF} & UART (\texttt{simpleuart}) &
    Serial output for UART benchmark reporting at 115200\,baud \\
  \bottomrule
  \end{tabular}
  \caption{System memory map of the implemented RISC-V SoC.}
  \label{tab:system-memory-map}
\end{table}

\begin{figure}[htbp]
  \centering
  \begin{tikzpicture}[
    font=\footnotesize,
    region/.style={draw, thick, minimum width=5.5cm, minimum height=1.1cm,
                   align=center},
    addr/.style={font=\ttfamily\footnotesize, anchor=east}
  ]
    \node[region, fill=blue!10]   (ram)  at (0, 3.3) {On-chip RAM \\ 64\,KB};
    \node[region, fill=orange!15] (mmio) at (0, 2.1) {SpMM accelerator \\ MMIO};
    \node[region, fill=green!12]  (gpio) at (0, 0.9) {GPIO \\ LEDs, switches, 7-seg};
    \node[region, fill=purple!12] (uart) at (0,-0.3) {UART \\ \texttt{simpleuart}};

    \node[addr] at (-3.0, 3.65) {0x0000\_0000};
    \node[addr] at (-3.0, 2.9) {0x0000\_FFFF};
    \node[addr] at (-3.0, 2.45) {0x1000\_0000};
    \node[addr] at (-3.0, 1.75) {0x1000\_00FF};
    \node[addr] at (-3.0, 1.27) {0x2000\_0000};
    \node[addr] at (-3.0, 0.55) {0x2000\_000F};
    \node[addr] at (-3.0, 0.1) {0x2000\_0100};
    \node[addr] at (-3.0,-0.70) {0x2000\_01FF};
  \end{tikzpicture}
  \caption{System memory map as a vertical address strip. The
           interconnect decodes each region using simple address
           comparisons.}
  \label{fig:memory-map}
\end{figure}

\section{Accelerator Programming Interface}

The accelerator is configured through the MMIO registers in
Table~\ref{tab:mmio-regs}. Firmware writes the dimensions and base
addresses, pulses \texttt{start}, polls \texttt{done}, and then reads the
output from shared RAM. There is no DMA engine, interrupt controller, or
driver layer.

\begin{table}[htbp]
  \centering
  \small
  \begin{tabular}{llp{10cm}}
  \toprule
  Offset & Name & Description \\
  \midrule
  \texttt{0x00} & \texttt{CTRL} &
    Control register: bit 0 = \texttt{start},
    bit 1 = \texttt{clear\_done}, bit 2 = \texttt{irq\_en}, bit 3 = \texttt{relu\_en},
    bit 4 = \texttt{int8\_mode} \\[4pt]
  \texttt{0x04} & \texttt{STATUS} &
    Status register: bit 0 = \texttt{busy}, bit 1 = \texttt{done} \\[4pt]
  \texttt{0x08} & \texttt{M} & Number of rows in $A$ and $C$ \\
  \texttt{0x0C} & \texttt{N} & Number of columns in $B$ and $C$ (dense width) \\
  \texttt{0x10} & \texttt{K} & Number of columns in $A$ / rows in $B$ \\
  \texttt{0x14} & \texttt{A\_VAL\_BASE} & Base address of CSR \texttt{values} array \\
  \texttt{0x18} & \texttt{A\_ROW\_BASE} & Base address of CSR \texttt{rowptr} array \\
  \texttt{0x1C} & \texttt{A\_COL\_BASE} & Base address of CSR \texttt{colidx} array \\
  \texttt{0x20} & \texttt{B\_BASE} & Base address of dense matrix $B$ (row-major) \\
  \texttt{0x24} & \texttt{C\_BASE} & Base address of output matrix $C$ (row-major) \\
  \texttt{0x28} & \texttt{NNZ} & Total number of nonzeros in $A$ \\
  \bottomrule
  \end{tabular}
  \caption{SpMM accelerator MMIO register map. All registers are 32-bit
           word-aligned. Base addresses are byte addresses into the shared
           64\,KB RAM.}
  \label{tab:mmio-regs}
\end{table}


\section{System Operation}

Figure~\ref{fig:system-sequence} shows the handshake between firmware and
accelerator for a single SpMM run. The CPU stages all inputs in BRAM, writes
the MMIO configuration, pulses \texttt{start}, and then polls the status
register. The accelerator runs asynchronously on Port~B of the BRAM until it
has written the last element of $C$, at which point it raises \texttt{done}.
The CPU clears the flag and proceeds to read $C$.

\begin{figure}[htbp]
  \centering
  \begin{tikzpicture}[
    font=\footnotesize,
    lane/.style={draw, thick, minimum width=5cm, minimum height=0.7cm,
                 align=center},
    seqbox/.style={draw, thick, rounded corners=2pt, fill=gray!10,
                   minimum width=4.4cm, minimum height=0.7cm, align=center},
    cpubox/.style={seqbox, fill=blue!10},
    accbox/.style={seqbox, fill=orange!15},
    arr/.style={-{Stealth[length=2mm]}, thick}
  ]
    \node[lane, fill=blue!5]   (cpuhdr) at (-3.2, 6.0) {\textbf{CPU (PicoRV32)}};
    \node[lane, fill=orange!5] (acchdr) at ( 3.2, 6.0) {\textbf{Accelerator}};

    \node[cpubox] (s1) at (-3.2, 5.0) {1. Stage CSR, $B$ in BRAM};
    \node[cpubox] (s2) at (-3.2, 4.1) {2. Set MMIO config registers};
    \node[cpubox] (s3) at (-3.2, 3.2) {3. Pulse \texttt{CTRL.start}};
    \node[cpubox] (s4) at (-3.2, 2.3) {6. Poll \texttt{STATUS.done}};
    \node[cpubox] (s5) at (-3.2, 1.3) {7. Write \texttt{CTRL.clear\_done}};
    \node[cpubox] (s6) at (-3.2, 0.3) {8. Read output matrix $C$};

    \node[accbox] (a1) at ( 3.2, 3.2) {4. Compute $C = A \times B$};
    \node[accbox] (a2) at ( 3.2, 2.3) {\quad\,\,tile over $N$, stream $A$};
    \node[accbox] (a3) at ( 3.2, 1.4) {5. Raise \texttt{STATUS.done}};

    \draw[arr] (s3.east) -- node[above, font=\scriptsize]{\texttt{start}} (a1.west);
    \draw[arr] (a3.west) -- node[above, font=\scriptsize]{\texttt{done}}  (s4.east);

    \draw[dashed] (-6.2, 5.55) -- (6.2, 5.55);
  \end{tikzpicture}
  \caption{System operation sequence for a single SpMM run. Steps 1--4 and
           7--8 execute on the CPU; steps 5--6 execute on the accelerator
           concurrently with the CPU's polling loop. The two lanes
           synchronise only through the \texttt{start} pulse and the
           \texttt{done} flag.}
  \label{fig:system-sequence}
\end{figure}

On a start pulse, the accelerator latches the dimensions and base
addresses, traverses $A$ in CSR form, reads $B$ from shared RAM, computes
$C$ tile by tile, optionally applies ReLU on writeback, and asserts
\texttt{STATUS.done}. In INT8 mode, the \texttt{values} array and $B$
matrix are interpreted as packed signed bytes and sign-extended before
multiplication.

\section{Design Constraints and Implications}

Four constraints shaped the platform. First, all benchmark data lives in
64\,KB of on-chip BRAM, which removes external-memory latency but limits
practical problem size to roughly $M = K = 64$ and $N = 32$. Second, the
accelerator must be driven from bare-metal firmware with no operating
system, DMA engine, or interrupt controller, so a small polling-based
MMIO interface was used. Third, the Cyclone~V DSP, BRAM, and routing
budgets limit the practical tile width; in this design $TN = 8$ was the
largest width that met timing reliably. Finally, all benchmarks must be
reproducible, so the firmware uses fixed seeds, element-wise software
checking, and PicoRV32's \texttt{rdcycle} counter.
