\chapter{Firmware, Benchmarking and Evaluation}
\label{ch:firmware}
\label{ch:results}

This chapter is the project's experimental half. It describes the
bare-metal firmware that drives the accelerator, the benchmark suite
that exercises it, and the measured results produced on hardware. The
order follows the order of the work: the firmware was built first, the
benchmark cases were designed once the firmware was stable, and the
cycle-count and correctness numbers reported at the end were produced
by running that firmware and that suite on the DE1-SoC.

\section{Firmware Benchmark Harness}
\label{sec:harness}

The bare-metal firmware drives the entire benchmark loop: matrix
generation, data preparation, accelerator configuration, timing,
correctness checking, and result reporting. It runs directly on
PicoRV32 from address \texttt{0x0000\_0000} with no operating system.
A short RISC-V assembly bootstrap sets the stack pointer to the top of
the 64\,KB RAM, clears the BSS section, and enters \texttt{main()},
which iterates over the 12 test cases and, for each, drives matrix
generation, the software reference, the accelerator launch, correctness
comparison, and UART reporting. The driver wraps the MMIO register
interface behind configuration, start, and polling functions; libc
stubs (\texttt{memset}, \texttt{memcpy}, \texttt{memmove}) cover the
minimal C library needs. The linker script maps all code and data
into the same 64\,KB RAM, with the stack growing downward from the top.

Matrix buffers are allocated statically below the stack, so the CSR
arrays, dense matrix $B$, and the output matrices (\texttt{c\_accel},
\texttt{c\_cpu}) occupy contiguous regions. The largest benchmark
(T6: $64 \times 64 \times 32$, 75\% sparsity, NNZ~$\approx 1024$)
consumes roughly
\[
  \underbrace{65 \times 4}_{\texttt{rowptr}} +
  \underbrace{1024 \times 4}_{\texttt{colidx}} +
  \underbrace{1024 \times 4}_{\texttt{values}} +
  \underbrace{64 \times 32 \times 4}_{B} +
  \underbrace{64 \times 32 \times 4}_{C} \approx 29\,\text{KB}
\]
of matrix data, well within the 64\,KB budget for firmware code and stack.

\subsection{Runtime data preparation}
\label{sec:data-prep}

Dense matrix $B$ is generated by a linear congruential PRNG seeded
from the benchmark case's \texttt{seed} field, with each entry sampled
as a signed integer in $[-127, 127]$; this range is directly usable as
INT32 in the full-precision path and as the pre-quantisation domain
for the INT8 path. A fixed seed makes $B$ deterministic across runs
and firmware revisions, which is what allows cycle-count regression
comparison across optimisation stages.

Sparse matrix $A$ is built in three phases. Nonzero positions are
placed first according to the selected pattern: a uniform pattern
gives each row $i$ a target count $n_i = \lfloor K(1 - s/100) \rfloor$
with column indices sampled without replacement; a row-skewed pattern
concentrates roughly 30\% of the NNZ budget in a minority of rows,
simulating social-graph hub nodes; and a clustered pattern biases
nonzeros toward rectangular blocks, mimicking the blocked sparsity
seen in finite-element problems. Each nonzero is then assigned a
random signed value in $[-127, 127]$. The CSR representation is built
last: column indices within each row are sorted ascending (required
for determinism and for matching the CPU reference), \texttt{rowptr}
is a prefix sum of per-row counts, and the sorted indices and values
are written to \texttt{colidx} and \texttt{values}. The sort also
makes the CSR form canonical across runs, which simplifies manual
inspection of accelerator output.

\subsection{INT8 quantisation and packing}
\label{sec:quant-flow}

Before an INT8 run, firmware applies symmetric runtime quantisation
separately to the CSR values array and to $B$. For a vector
$\mathbf{v}$ of length $L$, it computes $v_{\max} = \max_i |v_i|$,
sets $s = v_{\max}/127$, and quantises each element as
$q_i = \mathrm{clip}(\mathrm{round}(v_i/s), -127, 127)$. When
$v_{\max} = 0$, all outputs are zero and $s = 1$. This matches
Equation~\ref{eq:quant} from Section~2.6. The scale factor is
discarded immediately: because the correctness check compares
accelerator output against the \emph{quantised} CPU reference, $s$
need never be stored, which keeps both the accumulator logic and the
verification path scale-free.

After quantisation, four consecutive INT8 values are packed into a
single 32-bit word in little-endian byte order:
\begin{verbatim}
for (i = 0; i < len; i += 4) {
    word = ((uint32_t)(uint8_t)q[i+0])        |
           ((uint32_t)(uint8_t)q[i+1] << 8)   |
           ((uint32_t)(uint8_t)q[i+2] << 16)  |
           ((uint32_t)(uint8_t)q[i+3] << 24);
    dst[i/4] = word;
}
\end{verbatim}

$B$ is packed row by row, so a single 32-bit RAM read delivers four
aligned INT8 values that populate four lanes of the $B$-segment buffer
after sign extension, as described in Section~\ref{sec:fsm}.

\subsection{Software baseline}

The software reference is deliberately written in the simplest
possible form. The full-precision inner structure is:
\begin{verbatim}
for (row = 0; row < M; row++) {
    for (p = rowptr[row]; p < rowptr[row+1]; p++) {
        col = colidx[p];
        a   = values[p];
        for (n = 0; n < N; n++)
            C[row*N + n] += a * B[col*N + n];
    }
}
\end{verbatim}
\noindent\textbf{Listing~\ref{lst:spmm-cpu}}: PicoRV32 software SpMM
(full precision).
\label{lst:spmm-cpu}

Only the M-extension \texttt{MUL} instruction is used; there is no
SIMD, no loop unrolling, and no software prefetching. The baseline is
deliberately simple: its role is not to maximise CPU throughput but to
provide a credible, readable reference against which the accelerator's
structural advantage can be measured. The INT8 reference
(\texttt{spmm\_cpu\_q8}) has the same structure but operates on signed
8-bit operands with INT32 accumulation, and is the version used for
correctness comparison against the INT8 accelerator.

\subsection{Accelerator launch and reporting}

The driver exposes three entry points: \texttt{accel\_configure()}
writes the dimension and base-address registers,
\texttt{accel\_start\_mode()} issues the control write with the
\texttt{start} pulse and the \texttt{relu\_en}/\texttt{int8\_mode}
bits, and \texttt{accel\_wait\_done()} polls \texttt{STATUS.done}. A
launch wrapper chains them into the sequence used on every benchmark
case: write \texttt{clear\_done}, configure, capture $t_0$, issue
\texttt{start}, poll, capture $t_1$, write \texttt{clear\_done} again,
and report $t_1 - t_0$ as the accelerator cycle count.

Results are emitted as CSV over the on-chip UART at 115200\,baud
(50\,MHz / 434 cycles per bit); each line carries the test ID, $M$,
$K$, $N$, sparsity percentage and pattern, NNZ count, CPU and
accelerator cycle counts, derived speedup, and a PASS/FAIL verdict. A
condensed view is also driven onto the DE1-SoC's six seven-segment
displays, with the upper pair showing CPU cycles scaled by 64, the
middle pair raw accelerator cycles, and the lower pair the integer
speedup, so a running benchmark can be read at a glance without a host
terminal. The LEDs mirror the busy/done signals, lighting during
execution and clearing when computation completes.

Correctness is checked inline after each accelerator run by comparing
\texttt{c\_accel} against the INT8 CPU reference \texttt{c\_cpu}
element by element:
\begin{verbatim}
for (i = 0; i < M*N; i++) {
    if (c_accel[i] != c_cpu[i]) {
        pass = 0; break;
    }
}
\end{verbatim}
A full-precision CPU result is also computed in parallel but untimed,
and serves as a sanity reference for inspecting quantisation error.
Because CPU and accelerator apply the same symmetric INT8 quantisation
to the same data and accumulate identical INT32 partial sums, their
outputs must match exactly; any mismatch indicates a hardware bug
rather than a numerical artefact.


\section{Benchmark Suite}
\label{sec:bench-design}

The suite sweeps four independent axes so that any observed trend can
be interpreted analytically rather than anecdotally: matrix size, which
governs amortisation of fixed setup overhead; dense width $N$, which
sets the number of tile columns and the ratio of tile-management to
useful work; sparsity, which determines NNZ and therefore per-row
arithmetic load; and nonzero distribution pattern, which exposes
sensitivity to the spatial layout of NNZ. Table~\ref{tab:benchmark-inputs}
lists all 12 cases with the axis each primarily addresses. The same
cases are used unchanged across all three stages; this is what makes
stage-to-stage comparison meaningful.

\begin{table}[htbp]
  \centering
  \scriptsize
  \setlength{\tabcolsep}{9pt}
  \begin{tabular}{llrp{2.6cm}p{5.4cm}}
  \toprule
  Test ID & $M \times K \times N$ & Sparsity & Pattern & Purpose \\
  \midrule
  T1  & $8 \times 8 \times 8$    & 75\% & Uniform &
    Small debug/sanity case; establishes minimum speedup in the suite. \\
  T2  & $16 \times 16 \times 8$  & 75\% & Uniform &
    Small size-scaling point; 4$\times$ more arithmetic than T1. \\
  T3  & $32 \times 32 \times 8$  & 75\% & Uniform &
    Medium size-scaling point. \\
  T4  & $64 \times 64 \times 8$  & 75\% & Uniform &
    Largest square case; maximum $M$ for the current $M_{\max} = 64$. \\
  T5  & $64 \times 64 \times 16$ & 75\% & Uniform &
    Tests effect of doubling dense width relative to T4. \\
  T6  & $64 \times 64 \times 32$ & 75\% & Uniform &
    Tests effect of quadrupling dense width; 4 tile columns. \\
  T7  & $32 \times 32 \times 32$ & 50\% & Uniform &
    Low sparsity (dense); more nonzeros per row than T8/T9. \\
  T8  & $32 \times 32 \times 32$ & 75\% & Uniform &
    Mid-sparsity reference; matches T10/T11 except pattern. \\
  T9  & $32 \times 32 \times 32$ & 90\% & Uniform &
    Very sparse; exposes overhead-to-arithmetic ratio. \\
  T10 & $32 \times 32 \times 32$ & 75\% & Row-skewed NNZ &
    Same NNZ as T8; row imbalance sensitivity. \\
  T11 & $32 \times 32 \times 32$ & 75\% & Clustered NNZ &
    Same NNZ as T8; block locality sensitivity. \\
  T12 & $64 \times 64 \times 32$ & 90\% & Uniform &
    Large sparse stress case; wide output and few nonzeros. \\
  \bottomrule
  \end{tabular}
  \caption{Complete benchmark input set used across all recorded
           evaluation stages.}
  \label{tab:benchmark-inputs}
\end{table}

\vspace{2cm}

\section{Cycle-Count Definitions and Measurement Method}
\label{sec:cycle-defs}

CPU cycle counts bracket the timed software SpMM function only: the
outer row loop, CSR index reads, inner dense-column MAC loop, and the
reads and writes to $B$ and $C$. Matrix generation, quantisation, and
output-buffer clearing sit outside the bracket and are deliberately
excluded. Accelerator cycle counts bracket the complete offload call
and therefore include MMIO configuration, the \texttt{start} pulse,
hardware execution, and the final polling loop. This is an end-to-end
offload latency from the CPU's perspective, appropriate for
user-visible benefit but explicitly not a pure datapath-throughput
figure.

Timing uses the PicoRV32 \texttt{rdcycle} CSR, which returns the lower
32 bits of a hardware cycle counter clocked at 50\,MHz. The read is
wrapped in a short inline-assembly stub:
\begin{verbatim}
static inline uint32_t read_cycles(void) {
    uint32_t cyc;
    asm volatile("rdcycle %0" : "=r"(cyc));
    return cyc;
}
\end{verbatim}

The counter wraps at $2^{32}$ cycles, roughly 85 seconds at 50\,MHz.
The largest benchmark completes in under two million cycles, so
wrap-around is a non-issue and firmware needs no 32-to-64-bit
extension.

Speedup is defined as
\[
  \text{Speedup} = \frac{T_{\mathrm{CPU}}}{T_{\mathrm{ACCEL}}},
\]
with both times in CPU cycles on the same 50\,MHz clock. For the
RISC-V comparisons, $T_{\mathrm{CPU}}$ is the timed PicoRV32 software
result (INT32 for the baseline and DSP stages, INT8 for the INT8
stage). For the Cortex-M0 comparison, $T_{\mathrm{CPU}}$ is derived
from the processor's published CPI model applied to the same
benchmark definitions rather than measured on physical hardware;
Section~\ref{sec:cortex} discusses the implications.

\section{Baseline Accelerator Results}

Table~\ref{tab:baseline-results} gives the full baseline accelerator
results against the PicoRV32 software reference. Even unoptimised,
the accelerator delivers a consistent speedup, rising from
$12.12\times$ at T1 to $18.12\times$ at T4. T1 is the worst case
because its $8 \times 8$ matrices are too small to amortise the fixed
FSM overhead (the $M_{\mathrm{max}} = 64$ C-tile clear, the row loads,
and the final writeback), which dominates the cycle count at this
size. Section~\ref{sec:trend-analysis} analyses the amortisation
pattern across T1--T4.

\begin{table}[htbp]
  \centering
  \small
  \begin{tabular}{lp{4.2cm}rrrr}
  \toprule
  Test ID & $M \times K \times N$ / Sparsity & CPU cycles & Accel cycles & Speedup & Result \\
  \midrule
  T1  & $8 \times 8 \times 8$, 75\%    & 10,373   &    856 & $12.12\times$ & PASS \\
  T2  & $16 \times 16 \times 8$, 75\%  & 36,573   &  2,318 & $15.78\times$ & PASS \\
  T3  & $32 \times 32 \times 8$, 75\%  & 136,877  &  7,809 & $17.53\times$ & PASS \\
  T4  & $64 \times 64 \times 8$, 75\%  & 529,101  & 29,195 & $18.12\times$ & PASS \\
  T5  & $64 \times 64 \times 16$, 75\% & 986,829  & 58,112 & $16.98\times$ & PASS \\
  T6  & $64 \times 64 \times 32$, 75\% & 1,902,285 & 115,980 & $16.40\times$ & PASS \\
  T7  & $32 \times 32 \times 32$, 50\% & 951,213  & 58,112 & $16.37\times$ & PASS \\
  T8  & $32 \times 32 \times 32$, 75\% & 491,693  & 30,470 & $16.14\times$ & PASS \\
  T9  & $32 \times 32 \times 32$, 90\% & 215,263  & 13,844 & $15.55\times$ & PASS \\
  T10 & $32 \times 32 \times 32$, 75\%, row-skewed & 491,693 & 30,470 & $16.14\times$ & PASS \\
  T11 & $32 \times 32 \times 32$, 75\%, clustered  & 491,693 & 30,470 & $16.14\times$ & PASS \\
  T12 & $64 \times 64 \times 32$, 90\% & 800,155  & 49,663 & $16.11\times$ & PASS \\
  \bottomrule
  \end{tabular}
  \caption{Baseline PicoRV32 software versus baseline accelerator results.
           Speedup ranges from $12.12\times$ to $18.12\times$ across the
           12-case suite.}
  \label{tab:baseline-results}
\end{table}

\vspace{8cm}

\section{Parallel DSP MAC Enhancement Results}

Table~\ref{tab:dsp-results} gives the results after the DSP-oriented
MAC restructuring described in Chapter~\ref{ch:design}. Total
accelerator cycles fall from 427,299 to 324,568, a $1.32\times$
improvement, with every case improving over its baseline counterpart.
The gain is consistent but modest, and that is itself informative:
even with optimally scheduled DSP arithmetic, the accelerator is not
compute-bound. The $B$-segment memory transaction overhead ($TN = 8$
reads per nonzero at 32-bit precision) remains the dominant cost, and
no restructuring of the arithmetic alone can remove it.

\begin{table}[htbp]
  \centering
  \small
  \begin{tabular}{lrrrr}
  \toprule
  Test ID & CPU cycles & Accel cycles & Speedup & Result \\
  \midrule
  T1  & 10,373   &    737 & $14.07\times$ & PASS \\
  T2  & 36,573   &  1,859 & $19.67\times$ & PASS \\
  T3  & 136,877  &  6,024 & $22.72\times$ & PASS \\
  T4  & 529,101  & 22,021 & $24.03\times$ & PASS \\
  T5  & 986,829  & 43,781 & $22.54\times$ & PASS \\
  T6  & 1,902,285 & 87,301 & $21.79\times$ & PASS \\
  T7  & 951,213  & 43,781 & $21.73\times$ & PASS \\
  T8  & 491,693  & 23,296 & $21.11\times$ & PASS \\
  T9  & 215,263  & 10,988 & $19.59\times$ & PASS \\
  T10 & 491,693  & 23,296 & $21.11\times$ & PASS \\
  T11 & 491,693  & 23,296 & $21.11\times$ & PASS \\
  T12 & 800,155  & 38,188 & $20.95\times$ & PASS \\
  \bottomrule
  \end{tabular}
  \caption{Results after enabling the parallel DSP-oriented MAC
           enhancement. Total accelerator cycles fell from 427,299 to
           324,568, a $1.32\times$ reduction.}
  \label{tab:dsp-results}
\end{table}

\vspace{6cm}

\section{INT8 Packing Results}

Table~\ref{tab:int8-results} gives the final INT8 results, the largest
performance step in the project by a wide margin. Total accelerator
cycles fall from 324,568 to 148,737, a $2.18\times$ reduction. The
mechanism is exactly what the roofline analysis of
Chapter~\ref{ch:motivation} predicted: cutting $B$-segment reads per
nonzero from 8 to 2 reduces the dominant memory-bandwidth cost by a
factor of four on the reads themselves, and the overall cycle-count
impact is $2.18\times$ because not every cycle is memory-bound. Row
loads, nonzero index fetches, and the final C-tile writeback are
fixed overheads that do not shrink with INT8.

\begin{table}[htbp]
  \centering
  \small
  \begin{tabular}{lrrrr}
  \toprule
  Test ID & CPU cycles (INT8) & Accel cycles & Speedup & Result \\
  \midrule
  T1  & 10,735   &    567 & $18.93\times$ & PASS \\
  T2  & 38,087   &  1,111 & $34.28\times$ & PASS \\
  T3  & 142,999  &  2,964 & $48.25\times$ & PASS \\
  T4  & 553,655  &  9,747 & $\mathbf{56.80\times}$ & PASS \\
  T5  & 1,044,151 & 19,216 & $54.34\times$ & PASS \\
  T6  & 2,025,143 & 38,171 & $53.05\times$ & PASS \\
  T7  & 1,012,631 & 19,216 & $52.70\times$ & PASS \\
  T8  & 522,391  & 11,039 & $47.32\times$ & PASS \\
  T9  & 227,481  &  6,109 & $37.24\times$ & PASS \\
  T10 & 522,391  & 11,039 & $47.32\times$ & PASS \\
  T11 & 522,391  & 11,039 & $47.32\times$ & PASS \\
  T12 & 849,333  & 18,519 & $45.86\times$ & PASS \\
  \bottomrule
  \end{tabular}
  \caption{Final INT8-packed implementation results. Peak speedup of
           $56.80\times$ on T4; average speedup of $45.28\times$ across
           the 12-case suite. Total accelerator cycles fell from 324,568
           to 148,737, a $2.18\times$ reduction over the DSP stage.}
  \label{tab:int8-results}
\end{table}

\section{Stage-Level Summary}

Table~\ref{tab:stage-summary} consolidates the per-stage results into
a single view. All 12 benchmark cases pass their element-by-element
correctness checks at every stage, on hardware. Two observations stand
out. First, each stage improved on the previous one without a
correctness regression; this is a non-trivial outcome in a hardware
project where incremental FSM and datapath changes can readily
introduce off-by-one or sign-extension bugs. Second, the largest gain
came from reducing memory traffic (INT8 packing: $2.18\times$) rather
than from increasing arithmetic parallelism (DSP MAC: $1.32\times$),
directly confirming the memory-bandwidth bottleneck hypothesis
developed in Chapters~\ref{ch:motivation} and~\ref{ch:design}.

\begin{table}[htbp]
  \centering
  \small
  \begin{tabular}{lcccc}
  \toprule
  Stage & Avg speedup & Best case & Total accel cycles & Reduction vs prev. \\
  \midrule
  Baseline accelerator & $16.12\times$ & T4: $18.12\times$ & 427,299 & n/a \\
  Parallel DSP MAC     & $20.87\times$ & T4: $24.03\times$ & 324,568 & $1.32\times$ \\
  INT8 packed          & $45.28\times$ & T4: $56.80\times$ & 148,737 & $2.18\times$ \\
  \bottomrule
  \end{tabular}
  \caption{Summary of the three staged RISC-V accelerator evaluations
           across the full 12-case benchmark suite.}
  \label{tab:stage-summary}
\end{table}

\section{Comparison Against ARM Cortex-M0}
\label{sec:cortex}

Table~\ref{tab:cortex-results} compares the INT8 accelerator against a
software SpMM reference whose cycle counts come from an
instruction-level model of the ARM Cortex-M0 rather than a measurement
on physical hardware. The model applies the processor's published CPI
characteristics (predominantly single-cycle arithmetic, 2--3 cycle
word loads) to the same benchmark definitions used elsewhere. The
Cortex-M0 is included because it is a substantially more efficient
in-order baseline than PicoRV32 and is closer to the class of embedded
core with which an FPGA-attached SpMM block would plausibly be paired.
Speedups are correspondingly lower; the accelerator still reaches
$11.92\times$ on T4, a meaningful result for a small FPGA-attached
block. Being model-derived, these figures should be read as an
order-of-magnitude cross-check rather than a same-platform
measurement, and the PicoRV32 results remain the primary empirical
evidence of this chapter.

\begin{table}[htbp]
  \centering
  \small
  \begin{tabular}{lrrrr}
  \toprule
  Test ID & Cortex-M0 cycles & Accel cycles & Speedup & Result \\
  \midrule
  T1  & 2,952   &    567 & $5.21\times$  & PASS \\
  T2  & 9,448   &  1,111 & $8.50\times$  & PASS \\
  T3  & 31,912  &  2,964 & $10.77\times$ & PASS \\
  T4  & 116,200 &  9,747 & $\mathbf{11.92\times}$ & PASS \\
  T5  & 207,592 & 19,216 & $10.80\times$ & PASS \\
  T6  & 390,376 & 38,171 & $10.23\times$ & PASS \\
  T7  & 195,240 & 19,216 & $10.16\times$ & PASS \\
  T8  & 104,488 & 11,039 & $9.47\times$  & PASS \\
  T9  & 46,892  &  6,109 & $7.68\times$  & PASS \\
  T10 & 99,748  & 11,039 & $9.04\times$  & PASS \\
  T11 & 102,209 & 11,039 & $9.26\times$  & PASS \\
  T12 & 169,710 & 18,519 & $9.16\times$  & PASS \\
  \bottomrule
  \end{tabular}
  \caption{Model-based ARM Cortex-M0 software SpMM reference versus the
           final INT8 accelerator. The Cortex-M0 is a stronger baseline
           than PicoRV32, so speedups are correspondingly lower; the
           accelerator still achieves up to $11.92\times$ speedup.}
  \label{tab:cortex-results}
\end{table}

\section{Result Analysis}
\label{sec:trend-analysis}

\subsection{Scaling with problem size (T1--T4)}

Figure~\ref{fig:speedup-size} [placeholder] plots speedup against
matrix size for the three stages. Speedup rises consistently from T1
to T4 in all three, and in the INT8 stage grows from $18.93\times$ at
T1 ($8 \times 8 \times 8$) to $56.80\times$ at T4
($64 \times 64 \times 8$). The increase is sublinear in problem size:
an eight-fold increase in arithmetic from T1 to T2 yields only
$1.81\times$ more speedup, and the four-fold increase from T3 to T4
yields only $1.18\times$ more, indicating that the accelerator is
approaching the plateau where fixed overhead is a small fraction of
the total. The driver is the ratio of useful MAC work to FSM
overhead: T1 has roughly 12 nonzeros at 75\% sparsity, and the C-tile
clear alone (64 cycles for $M_{\max} = 64$) dominates, whereas T4's
1024 nonzeros spread the same fixed setup cost over far more useful
work.

\begin{figure}[htbp]
  \centering
  \begin{tikzpicture}
    \begin{axis}[
      width=0.82\linewidth,
      height=6.8cm,
      xlabel={Matrix dimension $M = K$},
      ylabel={Speedup vs PicoRV32 (INT32)},
      xmode=log, log basis x=2,
      xtick={8,16,32,64}, xticklabels={8,16,32,64},
      ymin=0, ymax=65,
      ymajorgrids, xmajorgrids,
      grid style={gray!25},
      legend pos=north west,
      legend cell align={left},
      legend style={font=\footnotesize, draw=gray!50},
      tick label style={font=\footnotesize},
      label style={font=\small},
      every mark/.append style={mark size=2.3pt},
    ]
      \addplot[mark=*, thick, blue!70!black]
        coordinates {(8,12.12) (16,15.78) (32,17.53) (64,18.12)};
      \addlegendentry{Baseline}
      \addplot[mark=square*, thick, red!70!black]
        coordinates {(8,14.07) (16,19.67) (32,22.72) (64,24.03)};
      \addlegendentry{Parallel DSP MAC}
      \addplot[mark=triangle*, thick, black!60!green]
        coordinates {(8,18.93) (16,34.28) (32,48.25) (64,56.80)};
      \addlegendentry{INT8 packed}
    \end{axis}
  \end{tikzpicture}
  \caption{Speedup versus matrix size for tests T1--T4 (fixed $N = 8$,
           75\% uniform sparsity) across the three optimisation stages.
           All three curves rise as problem size grows, confirming that
           fixed setup overhead (tile clear, row-bound load, final
           writeback) is amortised effectively at larger sizes. The
           baseline and DSP curves flatten by T4, whereas the INT8
           curve is still climbing steeply, indicating that INT8 packing
           not only shifts the absolute speedup upward but also extends
           the useful scaling regime.}
  \label{fig:speedup-size}
\end{figure}

\subsection{Effect of dense width (T4--T6)}

The sequence T4 ($N = 8$), T5 ($N = 16$), T6 ($N = 32$) tests how
speedup changes as more tile columns are processed at fixed $M = K$.
In the INT8 stage speedup moves from $56.80\times$ at T4 through
$54.34\times$ at T5 to $53.05\times$ at T6, falling only modestly as
$N$ grows. The accelerator handles multiple tile columns by repeating
the row traversal per column, separated by tile-clear and writeback
phases, and the CPU's inner loop also scales linearly with $N$; the
near-constant speedup therefore reflects linear scaling on both sides
and a preserved structural advantage. The mild decline reflects the
relative cost of the tile-management phases: at four tile columns,
four separate clear/writeback pairs run sequentially, and this fixed
per-column overhead accumulates on the accelerator's side but not the
CPU's.

\subsection{Effect of sparsity (T7--T9)}

The sequence T7 (50\%), T8 (75\%), T9 (90\%) fixes $M = K = N = 32$
and sweeps sparsity, giving $52.70\times \to 47.32\times \to
37.24\times$ in the INT8 stage. Speedup falls as the matrix becomes
sparser because the accelerator's per-row overhead (tile clear,
row-bound load) is paid regardless of row population, whereas the CPU
simply skips empty rows. At 90\% sparsity most rows of $A$ are empty
yet the accelerator still visits each of the 32, contributing fixed
traversal cycles. Even so, $37.24\times$ at T9 is a substantial
speedup and shows the design remains practical well into the sparsity
regimes typical of GNN workloads. Real social-graph GNN matrices
often exceed 99\% sparsity, where absolute cycle counts are low and
the relative speedup may fall further, but the acceleration in
absolute time would still be valuable. Figure~\ref{fig:speedup-sparsity}
plots the three-stage trend.

\begin{figure}[htbp]
  \centering
  \begin{tikzpicture}
    \begin{axis}[
      width=0.82\linewidth,
      height=6.8cm,
      xlabel={Sparsity (\%)},
      ylabel={Speedup vs PicoRV32},
      xtick={50,75,90},
      ymin=0, ymax=60,
      ymajorgrids, xmajorgrids,
      grid style={gray!25},
      legend pos=north east,
      legend cell align={left},
      legend style={font=\footnotesize, draw=gray!50},
      tick label style={font=\footnotesize},
      label style={font=\small},
      every mark/.append style={mark size=2.3pt},
    ]
      \addplot[mark=*, thick, blue!70!black]
        coordinates {(50,16.37) (75,16.14) (90,15.55)};
      \addlegendentry{Baseline}
      \addplot[mark=square*, thick, red!70!black]
        coordinates {(50,21.73) (75,21.11) (90,19.59)};
      \addlegendentry{Parallel DSP MAC}
      \addplot[mark=triangle*, thick, black!60!green]
        coordinates {(50,52.70) (75,47.32) (90,37.24)};
      \addlegendentry{INT8 packed}
    \end{axis}
  \end{tikzpicture}
  \caption{Speedup versus sparsity for tests T7--T9 (fixed
           $M = K = N = 32$, uniform pattern) across the three
           optimisation stages. The baseline and DSP stages are almost
           insensitive to sparsity because their cycle counts are
           dominated by per-row overhead that scales with $M$, not with
           NNZ. The INT8 stage shows a pronounced decline, as reducing
           memory-bandwidth cost makes the residual fixed-per-row
           traversal overhead a larger relative share of the
           accelerator's total cycles.}
  \label{fig:speedup-sparsity}
\end{figure}

\subsection{Pattern sensitivity (T8, T10, T11)}

T8, T10, and T11 share dimensions, sparsity, and NNZ, differing only
in how nonzeros are distributed (uniform, row-skewed, clustered). In
the INT8 stage all three report identical accelerator cycle counts of
11,039 and CPU counts of 522,391. This insensitivity is not
accidental: the accelerator processes each row sequentially regardless
of its nonzero count, and total NNZ, which sets the number of
\texttt{NZ\_FETCH} and \texttt{MAC} cycles, is held fixed across the
three patterns. Cycle count is therefore dominated by total NNZ rather
than by its spatial distribution. At larger scales, load imbalance
could become visible, since very dense rows would force many
sequential $B$-segment loads before the next row, but this effect is
not triggered at the present benchmark sizes. The Cortex-M0 model
shows small but nonzero differences (104,488, 99,748, and 102,209
cycles) consistent with pattern-dependent variation in CSR-index
reads and inner-loop iterations; a deeper analysis of the cause is
left for future work.

\section{Resource and Timing Evaluation}
\label{sec:resources}

Table~\ref{tab:quartus-final} gives estimated resource utilisation for
the three build variants. Final Quartus post-fit reports were still
being produced at the time of writing; the figures below are
pre-place-and-route estimates from the Quartus Fitter combined with
known design parameters, and are expected to be replaced by exact
post-fit numbers in the final revision. Utilisation is modest: roughly
16\% of ALMs, 5\% of M10K BRAMs, and 9\% of DSP blocks. The M10K BRAM
count is fixed at 16 across all three stages because it is determined
entirely by the shared 64\,KB RAM, not by the accelerator's internal
buffers, which sit in fabric registers; the DSP count of 8 matches the
8 parallel MAC lanes of the $TN = 8$ datapath. All three builds meet
the 50\,MHz timing constraint reported by the Quartus timing analyser.
The observed headroom on the Cyclone~V is a factual property of the
current build; concrete plans for using it (for example, a wider tile)
are deferred to Chapter~\ref{ch:conclusion}.

\begin{table}[htbp]
  \centering
  \small
  \begin{tabular}{lccccc}
  \toprule
  Build & ALMs & Registers & M10K BRAMs & DSP blocks & Fmax (MHz) \\
  \midrule
  Baseline accelerator  & $\approx$4,800 & $\approx$7,500 & 16 & 8 & 50+ \\
  Parallel DSP MAC      & $\approx$5,000 & $\approx$7,800 & 16 & 8 & 50+ \\
  INT8 packed (current) & $\approx$5,200 & $\approx$8,000 & 16 & 8 & 50+ \\
  \midrule
  Available (Cyclone~V) & 32,070 & 128,280 & 397 & 87 & n/a \\
  \bottomrule
  \end{tabular}
  \caption{Estimated resource utilisation for the three build variants on
           the Cyclone~V 5CSEMA5F31C6. All three designs meet the 50\,MHz
           timing constraint. Final post-fit numbers to be updated in the
           next revision.}
  \label{tab:quartus-final}
\end{table}

\section{Fairness, Limitations, and Threats to Validity}
\label{sec:threats}

The comparison is fair in several respects. CPU and accelerator
operate on identical inputs from the same fixed seeds, share the same
64\,KB on-chip RAM at the same 50\,MHz clock, and are checked against
the software reference before any cycle count is admitted. The
PicoRV32 baseline uses the M-extension \texttt{MUL} for its inner
multiplies, so it is not artificially slowed by emulation, and the
Cortex-M0 figures come from an instruction-level model applied to the
same benchmark definitions rather than a same-platform measurement.

Several limitations bound the claims that can be drawn. The PicoRV32
baseline is a simple in-order core with no cache, no speculation, and
no SIMD; against a modern superscalar CPU with vector extensions the
measured speedup would be considerably smaller. A more aggressively
optimised software implementation (software-pipelined inner loops or
explicit $B$-row reuse, for instance) would also narrow the gap. The
baseline's goal is credibility and simplicity, not maximum CPU
throughput.

As noted in Section~\ref{sec:cortex}, the Cortex-M0 comparison is
model-based. The model captures the coarse CPI behaviour of a
single-issue in-order Cortex-M0 reasonably well for the straight-line
arithmetic that dominates CSR SpMM, but it does not capture bus
contention, memory-subsystem effects, or compiler-specific code
generation on a real part. The Cortex-M0 figures are therefore an
order-of-magnitude cross-check, not a like-for-like measurement.

Each reported cycle count is a single recorded value rather than a
statistical distribution; counts for deterministic hardware are
repeatable, but a min/median/max summary would add confidence. The
test matrices are PRNG-generated rather than drawn from real GNN or
recommender-system datasets, so although they cover realistic sparsity
levels and pattern structures, behaviour on real workloads may differ.
All data lives in the 64\,KB on-chip BRAM; real SpMM workloads
typically span gigabytes and incur DRAM latency, which would affect
CPU and accelerator in different proportions, so the present results
reflect the best case of tightly coupled on-chip memory. Accelerator
cycle counts include MMIO configuration and polling overhead, which
is non-trivial on the smallest cases. Finally, the Quartus post-fit
resource and timing reports were still being finalised at the time of
writing; the design has been verified to close timing at 50\,MHz, but
exact ALM, BRAM, and DSP counts across all three variants remain
pending. These limitations bound the quantitative claims of the
chapter.

\section{Critical Evaluation}

The three-stage experiment gives a clear, quantitative demonstration
that the dominant bottleneck in CSR SpMM on this platform is memory
bandwidth, not arithmetic throughput. The $1.32\times$ gain from the
DSP MAC (an arithmetic improvement), set against the $2.18\times$
gain from INT8 packing (a memory-traffic improvement), is direct
experimental evidence of this. The result is consistent with the
analysis in Section~2.2: the operational intensity of CSR SpMM with
$TN = 8$ 32-bit operands is low, placing the design below the
roofline's compute--bandwidth crossover, and INT8 packing shifts
intensity upward by reducing operand fetch traffic, moving the design
toward the compute-bound region.

Direct comparison of the $56.80\times$ peak against
OuterSPACE~\citep{outerspace2018}, SpArch~\citep{sparch2020}, or
SIGMA~\citep{sigma2020} would not be meaningful: those systems target
orders-of-magnitude larger SpMM on dedicated memory hierarchies. The
relevant question is whether a tightly integrated, MMIO-driven FPGA
accelerator can deliver useful speedup on a constrained embedded
platform; the measured $56.80\times$ peak and $45.28\times$ average
over PicoRV32, and up to $11.92\times$ over the Cortex-M0 model,
answer in the affirmative. For wider context, Bell and
Garland~\citep{bell_garland2009} report $2$--$10\times$ GPU SpMV
speedups over scalar CPU code depending on matrix structure; the
present accelerator records higher speedups on a simpler SoC because
its software baseline is a slow in-order core and because INT8
packing sharply reduces the memory bottleneck.

Read alongside Section~\ref{sec:threats}, the results are internally
consistent and reproducible; correctness verification at every step
confirms that the recorded speedups reflect real computation rather
than artefacts of incorrect output; and the stage-to-stage trend is
explained by the same bandwidth argument that motivated the design.
The quantitative headline is simple: on this platform and against
these baselines, reducing operand traffic through INT8 packing was by
a wide margin the highest-leverage optimisation available.
